Experience Report: Haskell in Computational Biology

### 2012-06-28 mrfy_experience_report.tex
work on layout, formatting, and sections 2 and 3

\newif\ifnotcutting \notcuttingtrue % set if not cutting down to size

@@

\begin{document}

@@

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
The model gets its name because there are exactly
seven possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.

@@

To~be able to use Haskell, we had to reimplement the standard
algorithm for solving Viterbi's equations.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.
To~present some code,
we~begin with representations.

In~idiomatic Haskell, 
we might represent an individual amino acid~$x$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly,
and our hidden Markov models are stored using a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\begin{smallverbatim}
newtype AA = AA Int -- amino acid using HMMER coding
\end{smallverbatim}
Either way, a query sequence is (and should be) an immutable vector of
amino acids.
%type QuerySeq = Vector AA
%\end{smallverbatim}

The log of a probability is just a number, which we call a \texttt{Score}.
A transition probability connects two \texttt{StateLabel}s, and so we wrap it in
a value constructor:
\begin{smallverbatim}
newtype Score = Score Double

data TProb =
     TProb { logProbability :: Score
           , fromState :: StateLabel
           , toState :: StateLabel
           }

\end{smallverbatim}
We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed oval in \figref{plan7}) and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{edge}
function:
\smallverbatiminput{edge}
In MRFy, the first two arguments to \texttt{edge} are known at compile
time, and we expect \texttt{edge} to be inlined.

The representation of a node includes the probabilities of transitions
into the states of that node, plus the tables
of emission probabilities for the match and insertion states.
\remark{Why strictness annotations?}
\smallverbatiminput{hmmnode}

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show our code, we have to confess to a 
transformation of Viterbi's equations.
The~logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
And~to~make this file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}.
The transformed equations minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
A~negated log probability is called a~\emph{score}.

MRFy computes not only the minimum score but
also the path of states which produces that score.
Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallverbatiminput{vscore}
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

\tuningbox{
We~also made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}, so
\begin{smallverbatim}
fmap :: (a -> b) -> (Scored a -> Scored b)
\end{smallverbatim}
applies a function to the thing inside the container, without touching
the score.
And containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.
}

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
The other cases (\texttt{Ins} and \texttt{Del}) look similar.
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+,
so
the call to~\verb+vee''+ produces the same result as a recursive call
to~\verb+vee'+, but faster:
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

Memoization makes \verb+vee'+ perform just as well as the classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to improving performance.
By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.

>>>>

\newif\ifnotcutting \notcuttingfalse % set if not cutting down to size

@@

\begin{document}
\parskip=0.9\parskip % cheating

@@

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
The model gets its name because there are exactly
seven possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\remark{Problem: the subscripts are not so easy to read, and notation
is not quite consistent with \figref{viterbi}, definition of $a'$ at
bottom right.}
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure 
\ifnotcutting and evolutionary history \fi
with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
The~path that maximizes the
log of the probability is the most likely path.

@@

To~be able to use Haskell, we had to reimplement the standard
algorithm for solving Viterbi's equations.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.

Our code represents a query sequence as an immutable vector of amino
acids.
In~idiomatic Haskell, 
we might represent an individual amino acid~$x_i$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our models use a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\smallverbatiminput{aa}
The log of a probability is a negative number.
To~make these numbers more pleasant to read, we negate them,
which eliminates hordes of minus signs from our file formats.
The negated logarithm of a probability is called a \emph{score}.
\smallverbatiminput{score}
Type \texttt{Score} has a limited \texttt{Num} instance which permits
scores to be added and subtracted but not multiplied.

In a hidden Markov model, each transition is actually labeled with a
score, not a probability.
And we include the labels of the states that are
connected:\remark{\emph{Why} are the labels there?
And since they are apparently never used, why do they have strictness annotations???}
\begin{smallverbatim}
data TProb =   --- THIS DEFN IS A DAMN LIE
     TProb { logProbability :: Score
           , fromState :: StateLabel
           , toState :: StateLabel
           }
\end{smallverbatim}
In~a Plan7 hidden Markov model, as discussed above, 
there are seven transitions into each node:
\begin{smallverbatim}
data TProbs =  -- MORE DAMN LIES
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{edge}
function:\remark{Broken!  This function no longer appears in our examples}
\smallverbatiminput{edge}
In MRFy, the first two arguments to \texttt{edge} are known at compile
time, and we expect \texttt{edge} to be inlined.

Nodes are numbered, and the representation of a node includes 
the tables
of emission probabilities for the match and insertion states,
plus
the
probabilities of transitions 
into the states of that node.
\smallverbatiminput{hmmnode}

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
Our code implements a transformed version of Viterbi's equations, 
which operates on scores instead of on probabilities.
This version is
shown on the right-hand side of \figref{viterbi}.
THIS PART IS BROKEN.  THE A ARRAY NEEDS TWO SUBSCRIPTS, AND THE E
NOTATION IS INCONSISTENT (HAS MAGICALLY CHANGED FROM BRACKETS TO SUBSCRIPTS).
The transformed equations minimize the
score (i.e., the negated log probability) for each combination of
node~$j$, amino 
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallverbatiminput{vscore}
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

We~also made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}, so
\begin{smallverbatim}
fmap :: (a -> b) -> (Scored a -> Scored b)
\end{smallverbatim}
applies a function to the thing inside the container, without touching
the score.\remark{Changes in the code appear to have
made \texttt{fmap} irrelevant}
And containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

The \texttt{Scored} abstraction makes it easy to implement
Viterbi's equations.
For example, to compute $V_j^{\prime M}(i)$ using the equation at the top
right of \figref{viterbi}, we call 
\mbox{\texttt{vee\char`\'} \texttt{Mat} $j$ $i$}, 
which chooses the best (\texttt{minimum}) solution
from among the three predecessor states:
\smallverbatiminput{vfix}
In this code, EXPLAIN EXTEND, EXPLAIN ESCORE, EXPLAIN
ASCORE. 
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:\remark{BROKEN! \texttt{fmap} is gone}

Function~\verb+vee''+ is the \emph{memoized} version of~\verb+vee'+.
Calling~\verb+vee''+ produces the same result as calling~\verb+vee'+,
but faster: 
\smallverbatiminput{memo}
Functions \texttt{Memo.memo3} and \texttt{Memo.arrayRange} come from
Luke Palmer's
\path{Data.MemoCombinators} package.
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.

Memoization makes \verb+vee'+ perform just as well as the classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to dynamic programming.
By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.
